Using Monte Carlo Algorithms to Achieve Resiliency \& Performance at Scale for Linear and Nonlinear Solver Applications

### 2016-01-25 doc/ecp/pres.tex
Added slides for MCREX DOE presentation.

%%---------------------------------------------------------------------------%%
%% Shift Roadmap Presentation
%%---------------------------------------------------------------------------%%

\documentclass{beamer}
\usefonttheme[onlymath]{serif}
\usepackage{colortbl}
\usepackage{longtable,ltcaption,array}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{caption}
\usepackage[labelformat=empty]{subcaption}
\usepackage{tabularx}

@@

\setbeamercolor{item}{fg=MSUgreen}
\setbeamertemplate{headline}{}
\setbeamertemplate{navigation symbols}{}

@@

%%---------------------------------------------------------------------------%%
%% MATH STUFF
%%---------------------------------------------------------------------------%%

\newcommand{\Pn}{\textit{P}$\negthinspace_N$}
\newcommand{\SPn}{\textit{SP}$\negthinspace_N$}

\newcommand{\vOmega}{\ensuremath{\ve{\Omega}}}
\newcommand{\hOmega}{\ensuremath{\hat{\ve{\Omega}}}}

\newcommand{\sigs}{\ensuremath{\sigma_{\text{s}}}}
\newcommand{\sigf}{\ensuremath{\sigma_{\text{f}}}}
\newcommand{\sigsm}{\ensuremath{\sigma_{\text{s}m}}}
\newcommand{\sigsn}{\ensuremath{\sigma_{\text{s}n}}}

\newcommand{\phig}{\ensuremath{\Phi}}
\newcommand{\Sigmag}{\ensuremath{\mathbf{\Sigma}}}
\newcommand{\Dg}{\ensuremath{\mathbb{D}}}
\newcommand{\Fg}{\ensuremath{\mathbb{F}}}
\newcommand{\Qg}{\ensuremath{\mathbb{Q}}}
\newcommand{\ug}{\ensuremath{\mathbb{U}}}
\newcommand{\Ag}{\ensuremath{\mathbb{A}}}
\newcommand{\Bg}{\ensuremath{\mathbb{B}}}
\newcommand{\Cg}{\ensuremath{\mathbb{C}}}
\newcommand{\Jg}{\ensuremath{\mathbb{J}}}
\newcommand{\sg}{\ensuremath{\mathbb{S}}}

\newcommand{\aphig}{\ensuremath{\Phi^{\dagger}}}
\newcommand{\aSigmag}{\ensuremath{\mathbf{\Sigma}^{\dagger}}}

\newcommand{\apsi}[1]{\ensuremath{\psi^{\dagger\,#1}}}
\newcommand{\aphi}[1]{\ensuremath{\phi^{\dagger\,#1}}}
\newcommand{\aq}[1]{\ensuremath{q^{\dagger\,#1}}}
\newcommand{\ak}{\ensuremath{k^{\dagger}}}

\newcommand{\normal}{\ensuremath{\hat{\ve{n}}}}

% Phantom minus sign for helping with alignment of matrices
\newcommand{\phmin}{\ensuremath{\phantom{-}}}

\DeclareMathOperator{\diag}{diag}

@@

\title[Monte Carlo Linear Solvers]{Monte Carlo Linear Solvers}
\author[Slattery, Evans, Hamilton]{Stuart Slattery \\ Tom Evans
  \\ Steven Hamilton}
\date[11/4/14]{November 4, 2014}
\institute[ORNL]{\small Oak Ridge National Laboratory}
\titlegraphic{\includegraphics[width=1.5in]{ORNL_Balls.pdf}}

\defbeamertemplate{footline}{titlefoot}{
    \vspace{-0.5cm}
    \hspace*{0.25cm}
    \includegraphics[width=2cm]{doe_logo}
    \hfill
    \includegraphics[width=3.25cm]{WordMarkLeaf}
    \hspace*{0.5cm}
}

\setbeamertemplate{title page}
{
  \begin{tabular}{cr}
    \begin{minipage}{4.7cm}
      {\bf \Large\textcolor{MSUgreen}{\inserttitle}}\\

      \vspace{2\baselineskip}
      \insertauthor\\

      \insertinstitute\\

      \insertdate
    \end{minipage}
    &
    \begin{minipage}{5cm}
      \raggedright
      \vspace{-0.75cm}
      \includegraphics[height=9.8cm]{combined_background}
    \end{minipage}
  \end{tabular}
}

%%---------------------------------------------------------------------------%%
\begin{document}

@@

{
\setbeamertemplate{logo}{}
\setbeamertemplate{footline}[titlefoot]
\begin{frame}
\titlepage
\end{frame}
}

%%---------------------------------------------------------------------------%%
\begin{frame}{Acknowledgements}

  This material is based upon work supported by the U.S. Department of
  Energy, Office of Science, Advanced Scientific Computing Research
  program.

  \vfill

  This research used resources of the Oak Ridge Leadership Computing
  Facility at the Oak Ridge National Laboratory, which is supported by the
  Office of Science of the U.S. Department of Energy under Contract No.
  DE-AC05-00OR22725.

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Motivation}
  \vfill
\begin{itemize}
  \item As we move towards exascale computing, the rate of errors is expected
  to increase dramatically
  \vfil
  \begin{itemize}
  \item The probability that a compute node will fail during the course
    of a large scale calculation may be near 1
  \end{itemize}
  \vfill
\item Algorithms need to not only have increased concurrency/scalability
  but have the ability to recover from hardware faults
  \begin{itemize}
  \item Lightweight machines
  \item Heterogeneous machines
  \item Both characterized by low power and high concurrency
  \end{itemize}
  \vfill
\end{itemize}
\vfill
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Towards Exascale Concurrency and Resiliency}
  \begin{itemize}
    \item Two basic strategies:
      \vfill
      \begin{enumerate}
        \item State with current ``state of the art'' methods and make
          incremental modifications to improve scalability and fault
          tolerance
          \begin{itemize}
            \item Many efforts are heading in this direction, attempting
              to find additional concurrency to exploit
          \end{itemize}
        \vfill
        \item Start with methods having natural scalability and resiliency
          aspects and work at improving performance (e.g. Monte Carlo)
          \begin{itemize}
            \item Soft failures buried within the tally variance
            \item Hard failures mitigated by replication
            \item Concurrency enabled by several levels of parallelism
          \end{itemize}
      \end{enumerate}
  \end{itemize}
  \vfill
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Monte Carlo Methods

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo for Linear Systems}
  \begin{itemize}
    \item Suppose we want to solve $\mathbf{Ax}=\mathbf{b}$
    \vfill
    \item If $\rho(\mathbf{I-A})<1$, we can write the solution using the
      Neumann series
      \begin{equation*}
        \mathbf{x} = \sum_{n=0}^{\infty} (\mathbf{I-A})^n \mathbf{b}
         = \sum_{n=0}^{\infty} \mathbf{H}^n \mathbf{b}
      \end{equation*}
      where $\mathbf{H} \equiv ( \mathbf{I-A} )$ is the Richardson
      iteration matrix 
      \vfill
    \item Build the Neumann series stochastically
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Forward Monte Carlo}
\begin{itemize}
  \item Choose a row-stochastic matrix $\mathbf{P}$ and weight matrix
    $\mathbf{W}$ such that $\mathbf{H} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ij}| }
      {\sum_{j=1}^{N} | \mathbf{H}_{ij} | }
    \end{equation*}
  \item To compute solution component $\mathbf{x}_i$:
    \begin{itemize}
      \item Start a history in state $i$ (with initial weight of 1)
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_i$ based on current history weight
        and value of $\mathbf{b}_j$
    \end{itemize}
  \item A given random walk can only contribute to a single component of
    the solution vector
\end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Sampling Example (Forward Monte Carlo)}
  \begin{itemize}
    \item Suppose
  \begin{equation*}
    \mathbf{A} = \begin{bmatrix}
      \phmin 1.0 & -0.2 & -0.6 \\
      -0.4 & \phmin 1.0 & -0.4 \\
      -0.1 & -0.4 & \phmin 1.0 \end{bmatrix} \to
    \mathbf{H} \equiv (\mathbf{I-A}) = \begin{bmatrix}
       0.0 &  0.2 &  0.6 \\
       0.4 &  0.0 &  0.4 \\
       0.1 &  0.4 &  0.0 \end{bmatrix}
  \end{equation*}
    then
  \begin{equation*}
    \mathbf{P} = \begin{bmatrix}
       0.0 & 0.25 & 0.75 \\
       0.5 &  0.0 & 0.5 \\
       0.2 &  0.8 & 0.0 \end{bmatrix}, \quad
    \mathbf{W} = \begin{bmatrix}
       0.0 &  0.8 &  0.8 \\
       0.8 &  0.0 &  0.8 \\
       0.5 &  0.5 &  0.0 \end{bmatrix}
  \end{equation*}
    \vfill
    \item If a history is started in state $3$, there is a $20\%$ chance of
      it transitioning to state $1$ and an $80\%$ chance of moving to state
      $2$
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Solving the Heat Equation: Forward Method}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics<1>[width=4in]{direct_1.png}
      \includegraphics<2>[width=4in]{direct_10.png}
      \includegraphics<3>[width=4in]{direct_100.png}
      \includegraphics<4>[width=4in]{direct_1000.png}
    \end{center}
    \caption{
      \only<1>{\textbf{Forward solution.}
        \textit{\sn{2.5}{3} total histories.} }
      \only<2>{\textbf{Forward solution.}
        \textit{\sn{2.5}{4} total histories.} }
      \only<3>{\textbf{Forward solution.}
        \textit{\sn{2.5}{5} total histories.} }
      \only<4>{\textbf{Forward solution.}
        \textit{\sn{2.5}{6} total histories.} }
    }
  \end{figure}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Adjoint Monte Carlo}
\begin{itemize}
  \item Choose $\mathbf{P}$ and $\mathbf{W}$ such that
    $\mathbf{H}^{T} = \mathbf{P} \circ \mathbf{W}$
  \item Typical choice (Monte Carlo Almost-Optimal):
    \begin{equation*}
      \mathbf{P}_{ij} = \frac{| \mathbf{H}_{ji}| }
      {\sum_{i=1}^{N} | \mathbf{H}_{ji} |}
    \end{equation*}
  \item To estimate solution:
    \begin{itemize}
      \item Start a history in random state $i$ by sampling from distribution
        determined by $\mathbf{b}$
      \item Transition to new state $j$ based probabilities determined by
        $\mathbf{P}_i$
      \item Modify history weight based on corresponding entry in
        $\mathbf{W}_{ij}$
      \item Add contribution to $\mathbf{x}_j$ based on current history weight
    \end{itemize}
  \item A given random walk can contribute to many different components of
    the solution vector
\end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Solving the Heat Equation: Adjoint Method}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics<1>[width=4in]{adjoint_1.png}
      \includegraphics<2>[width=4in]{adjoint_10.png}
      \includegraphics<3>[width=4in]{adjoint_100.png}
      \includegraphics<4>[width=4in]{adjoint_1000.png}
      \includegraphics<5>[width=4in]{adjoint_10000.png}
      \includegraphics<6>[width=4in]{adjoint_100000.png}
      \includegraphics<7>[width=4in]{adjoint_1000000.png}
      \includegraphics<8>[width=4in]{adjoint_10000000.png}
    \end{center}
    \caption{
      \only<1>{\textbf{Adjoint solution.}
        \textit{\sn{1}{0} total histories.} }
      \only<2>{\textbf{Adjoint solution.}
        \textit{\sn{1}{1} total histories.} }
      \only<3>{\textbf{Adjoint solution.}
        \textit{\sn{1}{2} total histories.} }
      \only<4>{\textbf{Adjoint solution.}
        \textit{\sn{1}{3} total histories.} }
      \only<5>{\textbf{Adjoint solution.}
        \textit{\sn{1}{4} total histories.} }
      \only<6>{\textbf{Adjoint solution.}
        \textit{\sn{1}{5} total histories.} }
      \only<7>{\textbf{Adjoint solution.}
        \textit{\sn{1}{6} total histories.} }
      \only<8>{\textbf{Adjoint solution.}
        \textit{\sn{1}{7} total histories.} }
    }
  \end{figure}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Limitations of Monte Carlo}
\begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
    \begin{itemize}
      \item Central limit theorem is barrier to accurate solutions
    \end{itemize}
\end{itemize}
\begin{figure}
  \centering
  \includegraphics[width=3.5in]{Ifiss_ConvDiff}
\end{figure}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Residual Monte Carlo Methods

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Sequential Monte Carlo}
  \begin{itemize}
  \item Devised by John Halton in the 1960's as a residual method
\vfill
    \item Instead of directly solving $\mathbf{Ax}=\mathbf{b}$ with Monte
      Carlo, apply Monte Carlo to residual equation
      $\mathbf{A}\delta = \mathbf{r}$
    \vfill
    \item Preconditioned Richardson iteration using Monte Carlo as
      preconditioner:
      \begin{align*}
        \mathbf{r}^{k} &= \mathbf{b} - \mathbf{Ax}^{k} \\
        \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k} \\
        \mathbf{x}^{k+1} &= \mathbf{x}^k + \delta
      \end{align*}
    \vfill
    \item Exponential convergence achieved by decoupling Monte Carlo
      error from solution error
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Synthetic Acceleration}
  \begin{itemize}
  \item Devised by Evans and Mosher in the 2000's as an acceleration
    scheme for radiation diffusion problems (LANL)
    \vfill
  \item Can be abstracted as a general linear solver
    \vfill
  \item Combine with Richardson iteration as a ``smoother'' in between
    Monte Carlo steps:
    \begin{align*}
      \mathbf{r}^k &= \mathbf{b} - \mathbf{Ax}^k \\
      \mathbf{x}^{k+1/2} &= \mathbf{x}^k + \mathbf{r}^k \\
      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
      \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
    \end{align*}
  \end{itemize}
\end{frame}
%%---------------------------------------------------------------------------%%
\begin{frame}{Preconditioned MCSA}
\vspace*{-0.2in}
  {
    \small
    \[
    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
    \]
    \[
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
    \]
    \[
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
    \]
    \[
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
    \]
    \[
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
    \]
  }
  \vspace*{-0.1in}
  \begin{itemize}
    \item Requires explicit construction of preconditioned matrix
      \begin{itemize}
        \item This has generally limited preconditioner selection to diagonal,
          block diagonal, sparse approximate inverse approaches
      \end{itemize}
    \vfill
    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
      in $\mathbf{M^{-1}A}$
    \vfill
    \item Need to investigate preconditioning strategies that lead to sparse
      preconditioned systems (Michele Benzi, Emory)
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Residual Algorithm Comparison}
  \begin{figure}
    \includegraphics[width=3.5in]{seq_conv_100}
  \end{figure}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{3D Equilibrium Thermal Radiation Diffusion}
\vspace*{-0.1in}

{ \tiny
  T. Evans, S. Mosher, S. Slattery, S. Hamilton, ``A Monte Carlo
  synthetic-acceleration method for solving the thermal radiation
  diffusion equation,'' Journal of Computational Physics \textbf{258},
  pp.~338--358 (2014).  
}
\vspace*{-0.1in}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{figure}[ht!]
        \centerline{ \includegraphics[width=2in,clip]{mesh.pdf}}
      \end{figure}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{figure}[ht!]
        \centerline{ \includegraphics[width=2in,clip]{T_multi_mat.pdf}}
      \end{figure}
    \end{column}
  \end{columns}

\vspace*{-0.15in}
{\small
  \begin{table}
    \centering
    \begin{tabular}{cccc}
      \toprule
      \multirow{2}{*}{\bfseries Solver} &
      \bfseries Total &
      \bfseries Relative Time &
      \bfseries Relative Total \\
      & \bfseries Iterations &
      \bfseries Per Iteration &
      \bfseries Time \\
      \midrule
      CG (Jacobi) & 42,881 &  2.93 & 1.02 \\
      CG (ML) & 14,097 & 26.60 & 1.94 \\
      MCSA (Jacobi) & 90,129 & 1.0 & 1.0 \\
      \bottomrule
    \end{tabular}
  \end{table}
}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Path to Harder Problems}
  \begin{itemize}
    \item Experience with radiation diffusion suggests that for problems
      where spectral radius is low ($\lesssim 0.8$),
      MCSA can be competitive with leading methods
      \begin{itemize}
        \item Very low cost per iteration to approximate Neumann series
      \end{itemize}
    \vfill
    \item What about more challenging problems?
      \begin{itemize}
        \item Neumann series converges more slowly, so histories last longer
        \item More histories may be required
        \item What if Neumann series is not convergent?
      \end{itemize}
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}

  \center Polynomial Formulation

%%---------------------------------------------------------------------------%%
\begin{frame}{Stopping Criteria -- Weight Cutoff}
  \begin{itemize}
    \item Histories are usually terminated when the relative weight drops
      below a specified cutoff (i.e. $\frac{w}{w_0} < \tau$)
    \vfill
    \item Adjoint MCSA on JPWH\_991 matrix with diagonal preconditioning
      ($\rho ( \mathbf{I-D^{-1}A} ) \approx 0.98$):
  \end{itemize}
\begin{figure}
  \centering
  \includegraphics[width=2.5in]{WtCutoff}
\end{figure}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Weight vs. History Length Cutoff}
\begin{figure}
  \centering
  \includegraphics[width=3in]{WtVsLength}
\end{figure}
\begin{itemize}
  \item Unlike weight cutoff, history length cutoff ``saturates'' at
    some number of histories -- fixed length Neumann series has been
    effectively reproduced
  \item Saturation point is higher for longer history lengths
\end{itemize}
\end{frame}

>>>>

%%---------------------------------------------------------------------------%%
%% Beamer ORNL Presentation Template
%%---------------------------------------------------------------------------%%

\documentclass{beamer}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[compatibility=false]{caption}
\usepackage[labelformat=empty]{subcaption}
\usepackage{tikz}
\usepackage{ragged2e}

\usetikzlibrary{arrows}

@@

\setbeamercolor{item}{fg=MSUgreen}
\beamertemplatenavigationsymbolsempty
\setbeamertemplate{headline}{}
\setbeamertemplate{footline}{}
\setbeamertemplate{title page}
{
  \begin{tabular}{cr}
    \begin{minipage}{5.5cm}
      {\RaggedRight \bf \large \textcolor{MSUgreen}{\inserttitle}}\\
      \vspace{\baselineskip}\\
      \insertauthor\\
      \vspace{0.5\baselineskip}\\
      \insertinstitute\\
      \insertdate
    \end{minipage}
  \end{tabular}
}

@@

@@

\title[MCREX]{Using Monte Carlo Algorithms to Achieve Resiliency \& Performance
  at Scale for Linear and Nonlinear Solver Applications}
\author{Thomas Evans \and Steven Hamilton}
\date[DOE, 2016]{DOE RX-Solvers \\ January 29, 2016}
\institute[ORNL]{Oak Ridge National Lab}

%%---------------------------------------------------------------------------%%
\begin{document}
%%---------------------------------------------------------------------------%%

{
  \usebackgroundtemplate{
    \includegraphics[width=\paperwidth,height=\paperheight]{title_template}}
  \setbeamertemplate{logo}{}
  \begin{frame}
    \titlepage
  \end{frame}
}

%%---------------------------------------------------------------------------%%
\section*{Outline}
%%---------------------------------------------------------------------------%%

\begin{frame}
  \tableofcontents[pausesections]
\end{frame}

%%---------------------------------------------------------------------------%%
\section{Project Status}
%%---------------------------------------------------------------------------%%
\subsection{Project Goals and Target Applications}
%%---------------------------------------------------------------------------%%

\begin{frame}{Project Objectives}

  The MCREX project has 3 objectives for the development of Monte Carlo solver
  methods for linear systems:
  \begin{itemize}
  \item Analyze and improve the iterative performance of Monte Carlo
    algorithms for linear systems
  \item Analyze and characterize the resilient aspects of Monte Carlo
    algorithms for linear systems
  \item Develop and model highly scalable parallel algorithms for the
    implementation of Monte Carlo algorithms for linear systems
 \end{itemize}
 The end goal is to develop a set of resilient, scalable algorithms for the
 solution of linear systems.  We have made significant progress in each of
 these areas as summarized below.

@@

\begin{frame}{Target Applications}

  \begin{itemize}

  \item Models defined by PDEs that result in sparse linear systems
    \begin{itemize}
    \item fluid flow
    \item radiation transport
    \item electrostatics
    \end{itemize}

  \item We have begun to analyze eigenvalue systems that are important in
    network systems analysis and data mining

  \item Much of our analysis work has used and idealized
    advection-diffusion-reaction system that has properties common to these
    application areas

  \item Extension to non-linear Navier-Stokes dropped due to budget
    constraints of funded proposal
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\subsection{Overview and Status}
%%---------------------------------------------------------------------------%%

\begin{frame}{Monte Carlo for Linear Systems}
  \begin{itemize}
    \item Suppose we want to solve $\mathbf{Ax}=\mathbf{b}$
    \vfill
    \item If $\rho(\mathbf{I-A})<1$, we can write the solution using the
      Neumann series
      \begin{equation*}
        \mathbf{x} = \sum_{n=0}^{\infty} (\mathbf{I-A})^n \mathbf{b}
         = \sum_{n=0}^{\infty} \mathbf{H}^n \mathbf{b}
      \end{equation*}
      where $\mathbf{H} \equiv ( \mathbf{I-A} )$ is the Richardson
      iteration matrix
      \vfill
    \item Build the Neumann series stochastically
  \end{itemize}

@@

\begin{frame}{Solving the Heat Equation}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics<1>[width=4in]{../NCState_seminar_2014/adjoint_1}
      \includegraphics<2>[width=4in]{../NCState_seminar_2014/adjoint_10}
      \includegraphics<3>[width=4in]{../NCState_seminar_2014/adjoint_100}
      \includegraphics<4>[width=4in]{../NCState_seminar_2014/adjoint_1000}
      \includegraphics<5>[width=4in]{../NCState_seminar_2014/adjoint_10000}
      \includegraphics<6>[width=4in]{../NCState_seminar_2014/adjoint_100000}
      \includegraphics<7>[width=4in]{../NCState_seminar_2014/adjoint_1000000}
      \includegraphics<8>[width=4in]{../NCState_seminar_2014/adjoint_10000000}
    \end{center}
    \caption{
      \only<1>{\textbf{Adjoint solution.}
        \textit{\sn{1}{0} total histories.} }
      \only<2>{\textbf{Adjoint solution.}
        \textit{\sn{1}{1} total histories.} }
      \only<3>{\textbf{Adjoint solution.}
        \textit{\sn{1}{2} total histories.} }
      \only<4>{\textbf{Adjoint solution.}
        \textit{\sn{1}{3} total histories.} }
      \only<5>{\textbf{Adjoint solution.}
        \textit{\sn{1}{4} total histories.} }
      \only<6>{\textbf{Adjoint solution.}
        \textit{\sn{1}{5} total histories.} }
      \only<7>{\textbf{Adjoint solution.}
        \textit{\sn{1}{6} total histories.} }
      \only<8>{\textbf{Adjoint solution.}
        \textit{\sn{1}{7} total histories.} }
    }
  \end{figure}
\end{frame}

\begin{frame}{Limitations of Monte Carlo}
  \begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
    \begin{itemize}
    \item Central limit theorem is barrier to accurate solutions
    \end{itemize}
    \begin{figure}
      \centering
      \includegraphics[width=1.75in]{../NCState_seminar_2014/Ifiss_ConvDiff}
    \end{figure}
  \item There is a potentially restrictive limit on the spectral properties of
    $\mathbf{H}$
    \begin{equation}
      \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
      \frac{H_{ij}^2}{P_{ij}}
    \end{equation}
  \end{itemize}
\end{frame}

\begin{frame}{Monte Carlo Synthetic Acceleration}
  \begin{itemize}
  \item Address issue of convergence constrained by CLF
    \vfill
  \item Can be abstracted as a general linear solver
    \vfill
  \item Combine with Richardson iteration as a ``smoother'' in between
    Monte Carlo steps:
    \begin{align*}
      \mathbf{r}^k &= \mathbf{b} - \mathbf{Ax}^k \\
      \mathbf{x}^{k+1/2} &= \mathbf{x}^k + \mathbf{r}^k \\
      \mathbf{r}^{k+1/2} &= \mathbf{b} - \mathbf{Ax}^{k+1/2} \\
      \hat{\mathbf{A}} \mathbf{\delta} &= \mathbf{r}^{k+1/2} \\
      \mathbf{x}^{k+1} &= \mathbf{x}^{k+1/2} + \delta
    \end{align*}
  \item We still must address issue of convergence constraint by spectral radius
  \end{itemize}
\end{frame}

\begin{frame}{Preconditioned MCSA}
\vspace*{-0.2in}
  {
    \small
    \[
    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k})
    \]
    \[
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
    \]
    \[
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{x}^{k+1/2})
    \]
    \[
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
    \]
    \[
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
    \]
  }
  \vspace*{-0.1in}
  \begin{itemize}
    \item Requires explicit construction of preconditioned matrix
      \begin{itemize}
        \item This has generally limited preconditioner selection to diagonal,
          block diagonal, sparse approximate inverse approaches
      \end{itemize}
    \vfill
    \item Sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity
      in $\mathbf{M^{-1}A}$
    \vfill
    \item Need to investigate preconditioning strategies that lead to sparse
      preconditioned systems
  \end{itemize}
\end{frame}

\begin{frame}{Monte Carlo as Approximate Polynomial Preconditioning}
  \begin{itemize}
    \item Using history length cutoff, Monte Carlo process is approximating
      a fixed length Neumann series polynomial
      \[
      x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
      \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
      \]
    \vfill
    \item As number of histories grows, iteration count becomes identical to
      using ``true'' polynomial
    \vfill
    \item Why limit ourselves to the Neumann series polynomial?
      \begin{itemize}
        \item Chebyshev or GMRES polynomials are viable alternatives
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Neumann Series Polynomial}
  \begin{table}
    \caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
      Values are MCSA iteration counts (timing in milliseconds)
      \label{tab:lap_adjoint_neumann}}
    \centering
    \begin{tabular}{cccc}
      \toprule
      Histories per & \multicolumn{3}{c}{Polynomial Order} \\
      \cmidrule(lr){2-4}
      Iteration & 2 & 4 & 6 \\
      \midrule
      250 &  775(100) & 651(96) & 664(110) \\
      500 &  725(122) & 502(104) & 394(95) \\
      1000 & 707(174) & 482(179) & 366(144) \\
      2000 & 703(280) & 471(259) & 356(251) \\
      4000 & 698(494) & 467(497) & 350(458) \\
      8000 & 697(923) & 464(905) & 350(873) \\
      16000 & 695(1796) & 462(1768) & 347(1711) \\
      \bottomrule
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}{Chebyshev Polynomial}
  \begin{table}
    \caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
      Values are MCSA iteration counts (timing in milliseconds)
      \label{tab:lap_adjoint_cheby}}
    \centering
    \begin{tabular}{cccc}
      \toprule
      Histories per & \multicolumn{3}{c}{Polynomial Order} \\
      \cmidrule(lr){2-4}
      Iteration & 1 & 2 & 3 \\
      \midrule
      250 & - & - & - \\
      500 & - & - & - \\
      1000 & - & - & - \\
      2000 & - & 328(134) & - \\
      4000 & - & 296(210) & - \\
      8000 & - & 288(380) & 262(423) \\
      16000 & 1132(1866) & 283(721) & 175(550) \\
      \bottomrule
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}{Polynomial Methods -- Summary}
  \begin{itemize}
    \item Using history length cutoff rather than weight cutoff can lead to
      large reductions in iteration counts
      \vfill
    \item A good approximation to a few terms in the Neumann series
      performs better than a statistically noisy approximation to
      full series
      \vfill
    \item Significant reduction in iteration counts are possible using
      alternate polynomials, but generally outweighed by increase in number
      of histories required
      \begin{itemize}
        \item May be very beneficial from resiliency and parallel efficiency
          standpoints
        \item Will be re-evaluated in future efforts
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Sparse Approximate Inverse Preconditioning}

@@

%%---------------------------------------------------------------------------%%
\subsection{Software}
%%---------------------------------------------------------------------------%%

\begin{frame}{MCSA Implementations}

  \begin{itemize}
  \item {\bf MCLS} (\underline{github.com/sslattery/MCLS})
    \begin{itemize}
    \item Massively parallel (MPI) implementation of domain decomposed MCSA
      algorithm
    \item Implemented in Trilinos framework and callable through Trilinos'
      abstract numerical algorithms (Thyra, Stratimikos) interface
    \item Works with any Tpetra row matrix
    \end{itemize}
  \item {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
    \begin{itemize}
    \item GPU-accelerated on-node MC solve
    \item Domain-local MC solves implemented for distributed systems as an
      additive Schwartz method
    \end{itemize}
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\section{Transition and Roadmap}
%%---------------------------------------------------------------------------%%

\begin{frame}
\vfill
\centering
\huge Questions?
\vfill
\end{frame}
